PART TWO - CREEDS Drug Data
Goal: Add a column 'drugbank-id' to the automated CREEDS drug data.

1. Reformatting the DrugBank data
```{r}
# Read in drugbank data
drugbank <- read.csv("data/drugbank_vocabulary.csv")
# Create new data frame "drug_mapping" with subset of columns: Common.name, Synonyms, DrugBank ID
drug_mapping <- drugbank %>% select(Common.name, Synonyms, DrugBank.ID)
# Use separate_rows() to separate out the synonyms
drug_mapping <- separate_rows(drug_mapping, Synonyms, sep = "\\|")
```
# Need a way to remove all drugs with synonyms that are invalid UTFs !!

2. Reformatting the CREEDS data
```{r}
# Read in CREEDS data
# Genes: Manual and Auto
gene_pert_manual <- read.csv("data/single_gene_perturbations-v1.0.csv", stringsAsFactors = FALSE)
gene_pert_auto <- read.csv("data/single_gene_perturbations-p1.0.csv", stringsAsFactors = FALSE)

# Drugs: Manual and Auto
drug_pert_manual <- read.csv("data/single_drug_perturbations-v1.0.csv", stringsAsFactors = FALSE)
drug_pert_auto <- read.csv("data/single_drug_perturbations-p1.0.csv", stringsAsFactors = FALSE)

# Diseases: Manual and Auto
disease_pert_manual <- read.csv("data/disease_signatures-v1.0.csv", stringsAsFactors = FALSE)
disease_pert_auto <- read.csv("data/disease_signatures-p1.0.csv", stringsAsFactors = FALSE)

# Use separate_rows() on the CREEDS data to separate out multiple drug names.
  # I believe we are supposed to separate out both control IDs and pert IDs. (??)
gene_pert_auto_copy <- gene_pert_auto
gene_pert_auto_copy <- separate_rows(gene_pert_auto_copy, ctrl_ids, sep = "\\|")
gene_pert_auto_copy <- separate_rows(gene_pert_auto_copy, pert_ids, sep = "\\|")
```

3. Putting the data together
We combine CREEDS data with DrugBank data.
  Goal: To see which drugs in CREEDS data are in DrugBank data.
# Why do we want to use a left join vs a different join?
We use a left join to keep all of the observations in the CREEDS data
# What should we put in the "by" field? (drug_name or drugbank_id? why?)
drugbank_id
# What does the result look like?
It's just another data frame (CREEDSinDrugBank)
```{r}
# We do a left join() where x is CREEDS data and y is "drug_mapping"
CREEDSinDrugBank <- left_join(drug_pert_auto_copy, drugNames, by = 'drug_name')
```

4. Look at drug instance mapping to DrugBank
# Which drug instances do not map to DrugBank?
These drug instances were found in CREEDS data but not in the DrugBank.
In other words, these are novel drugs found by CREEDS.

  1) Look for NAs in the drugbank_id field.
```{r}
NADrugBankIDs <- copyCREEDS %>% filter(is.na(DrugBank.ID))
# I didn't use group_by() ; It seems useful but I couldn't get a way for it to select only the NAs in the drugbank_id field. (??)
dim(NADrugBankIDs) # 83012 DrugBank IDs which are NAs
```
  
  2) Make a list of the drug names for the drug instances without mapping.
```{r}
noIDdrugNames <- NADrugBankIDs$drug_name
View(noIDdrugNames)
```

  3) Write this list out to a file.
```{r}
# Table with information about drug names with no mapping and their frequencies.
noIDdrugNamesFreq <- table(noIDdrugNames)
View(noIDdrugNamesFreq)
# Write this list out to a file
write.csv(noIDdrugNamesFreq, "data/noIDdrugNamesFreq.csv")
```

5. Identify studies
We will do a thorough search for studies with >= 8 samples in the data.
We need studies that have >= 4 males and >= 4 females to use in our meta-analysis.

Once identified, we will sex-label these studies and check to see whether we can use them.
# Count the number of control samples in each study (studies with >= 8 samples)
    -> ("ctl_studies.txt")
```{r}
subset <- subset[!duplicated(subset),]
tab <- table(subset$geo_id)
val <- tab[tab >= 8]
write.csv(val, "ctl_studies.txt")
```
# Count the number of perturbed samples in each study with a particular drug.
    -> ("drug_studies.txt")
```{r}

write.csv(xxxxx, "drug_studies.txt")
```

6. Exploratory Analysis
Think about how to organize/examine the data.

How many controls are on average in a study? How many samples? (use npert, nctl for this).
# mean(val$n) results in 16.08 controls on average in a study. 
Which drugs are present?

How many instances of each drug?

Make a histogram of # drugs for each drug using geom_histogram from ggplot2
# Should show # of drugs vs. name of drug
```{r}
# Get the table of drugs and their frequencies in descending order.
new_data <- drugFreqs %>% arrange(desc(freq))
# Get the 15 drugs with the highest counts.
new_data <- new_data[1:15, ]
```

Look at all of this separated by organism


What other questions do you have about the data?

Make another histogram about some other aspect of the data.





