(Attempts to) label samples with DrugBank IDs

```{r}
require('tidyverse')
require('MetaIntegrator')
```

1. Extract unique columns
```{r}
gse28006 <- getGEOData("GSE28006")
pData28006 <- gse28006$originalData$GSE28006$pheno

# identify number of levels for each column
# why does this work? - make sure you understand
nlevels.per.col <- apply(pData28006, 2, function(x) nlevels(as.factor(x))) 
sample.specific.cols <- colnames(pData28006)[nlevels.per.col!= 1] 

# note: geo_accession and supplementary_file are not helpful
sample.descript.cols <- setdiff(sample.specific.cols, c("geo_accession"))
summary(pData28006[,sample.descript.cols])

# note - do you notice any misspellings? look at "description" column for this example. we have to be careful! it's human-annotated

# studies will have very variable numbers of columns

# now put this together into a single field - you will want to use "paste"
sample.labels <- apply(pData28006[,sample.descript.cols], 1, function(x) paste(x, collapse="\t"))
# what does this command do? how does it do this
#### It combines all of the information in the unique sample into one line, separated by tabs.

# using pData this because we want character version
# why are we using a "1"" here?
### To combine it via rows

# Write this to a file 
write.csv(sample.labels, "data/sample.labels.gse28006.csv")
```

2. Reformat DrugBank data
```{r}
# Read in the drugbank data you were working with from DrugLabels.Rmd
drugbank <- read.csv("data/drugbank_vocabulary.csv")

# Create a new data frame "drug_mapping" with a subset of the columns: Common.name, Synonyms, DrugBank ID
drug_mapping <- drugbank[, c("Common.name", "Synonyms", "DrugBank.ID")]

# Use separate_rows() to separate out the synonyms
drug_mapping_sep <- separate_rows(drug_mapping, Synonyms, sep = "\\|")

# Since we are searching for the synonyms or common names, we need to condense this into a data.frame that has two columns:'drug_name' and 'drugbank_id'. To do this, I would run something like:
df1 <- select(drugbank,c("Synonyms", "DrugBank.ID"))
df1 <- separate_rows(df1, Synonyms, sep="\\|")
df2 <- select(drugbank,c("Common.name", "DrugBank.ID"))

# then re-name the columns of the first data frame so they are the same
df1 <- rename(df1, Common.name=Synonyms)
df.combined <- rbind(df1[colnames(df2)], df2) # <<< ??? DISCUSS THIS >>>>
# then run `rbind()` to put the two together
```

```{r}
# remove any rows with empty values
df.combined <- df.combined[df.combined$Common.name != ""]

# trim all the strings
???
```


3. Search for drug names (toy example)
# Returns a data frame with the drugs and samples after searching all the sample labels for each drug
  # Gets a vector with TRUE / FALSE for each drug
    # Puts together the drugs and samples in a data frame
```{r}
# search in the list
# as far as I can tell, the easiest thing to do is to search -ALL- the sample labels for each drug
# at the end, you will get a vector with TRUE / FALSE for each drug
# these will be put together in a data.frame --> the columns are the drug, the rows are the samples


# step thru a toy example, see what it looks like
#  # c("a1c", "bb", "3a") are the sample labels, then:

# list of drugs
drugs <-  list("DB1"="1|3", "DB2"="b") 

# 
toy.drug.df <- sapply(drugs, function(x) str_detect(c("a1c", "bb", "3a"), x))  

# now label each "sample" with a drug
#  sample 1 and 3 should be labeled DB1, sample 2 should be labeled DB2
drug.idx <- apply(toy.drug.df, 1, function(x) which(x)) 
# why does this work? what are we doing? make sure you understand it
# look at the input, output -and- try it in pieces if you don't

# try making this more complicated - what if there are multiple drugs in a row? (try "b|c") no drugs (try "bc")?
# think about how will we have to address this
#  come up with a couple more test cases
#  make sure it works with at least 4 made-up "drugs""

# write a function to get the output formatted nicely
#  input: drug.idx list
#         list of samples
#  output: data.frame formatted with:
#             | sample.id | drugbank.id |
#  error handling:
#     - if there are no drugs, output an error message, and then the drugbank.id should contain "NA" in that part
#     - if there are multiple drugs, output an error message, and then the drugbank.id in the df will contain the drugbank IDs separated by an or "DB1 | DB2"
#     - make the error messages informative
```

Function that outputs DrugBank ID as another column at the end of the table of phenotypic data.


